Zohmg---a large scale data store for aggregated time-series-based data. [DRAFT]

### 2009-07-13 doc/report/msc-report.tex
minor changes to thesis.
thesis improvements.

Any organization will eventually want to analyze various characteristics of
its service and operations. The services can be anything from web sites to
theme park; the analysis might include characteristics such as the number of
visitors per day, or the average time a theme park guest spends queueing for a
particular ride.

Generally, the services and operations will generate a log for later analysis.
As the services grow in size, so will the data sets they generate.

An important property of these data sets is that they generally have a regular
structure and are broken down into records that are independent of each other.
The records can be analyzed one at a time without losing any of their meaning
and are as such very well suited for parallel analysis.

Each log record typically contains a number of fields. For example, a log of
visits to a web site might contain a timestamp, information about what country
the visitor is from and what web browser she used.

When the records in the data set have been analyzed, one can draw conclusions
about what countries generate the most traffic to the site, which web browser
is most popular in a certain country, etc. Summaries of data are generally
called aggregates. To aggregate data, for example to compute the sum of
visitors for all European countries, is a common technique for speeding up
data lookup in data stores.

The size of the data sets at Last.fm is in the range of several gigabytes per
day. Some data sets are many times larger. It is our belief that this is
typical.

@@

Organizations measure their operations by considering many variables. When
these variables are placed in a spreadsheet, which organizations are wont to
do, the variables are set on the x and y axes. Each axis represents a logical
grouping of variables. In the example of a log file of visits to a web site
above, country of origin is one such variable. A variable is generally called
a dimension in this thesis.

The International Aerodyne Corporation, a leading exporter of woodworking
nails, uses a spreadsheet to keep track of the number of nails exported to
each country every month (see table 1.1).

@@

% TODO: renskriv these thoughts.

Last.fm is an internet radio station focusing on content, being a musical
encyclopedia. Last.fm lets users submit what music they listen to, called
\textit{scrobbling}, this user submitted data is analyzed and Last.fm offer
their users personalized charts, music and event recommendations, based ther
listening. Last.fm also streams content, via internet, to users. The Last.fm web
site ties together all these areas and presents it to the users.

The web site, radio streaming and submissions (scrobbling) generates gigabytes
of data every day.

@@

% TODO: define aggregates!

@@

In addition, the secondary goal is to build a tool that allows users to easily
store their data sets in the data store. The tool should provide a simple – as
in minimalistic – mechanism for data analysis (import) and querying (export).

% is mapreduce part of the goal? hardly.
The data analysis will require the user to write a MapReduce map. The queries
will be performed via an export API for consumption by computer programs.

% these are more likely part of the scope.
The queries will always be made over a time range and for a single unit. The
tool will support analytical functions such as drill-down, slicing, etc.

The tool should be aware of dimensions in general and the dimension of time in
particular. Measurements will be specified in a unit. There will be no limit
on the number of units.

There should be no theoretical boundary on the number of dimensions. Common
queries must be answered expediently to allow for analysts to explore data in
real time.

\section*{Scope}
% Avgransning.

% TODO: argue why we do these avgränsningar.

The data store is just that: a data store with import and export
functionality. The data store does not interpret the data it store.

The data store is time-series-based: all data must include the time dimension.

There is no functionality for plotting data sets, etc.

\section*{Related work}

@@

BigTable is a data store designed for traversal of very large data sets.
BigTable's cells are basically indexed by row key, column key, and a time-stamp.
Each row has a key and any number of cells attached to it. The rows in a table
are sorted by their keys. The typical use-case of BigTable is to scan a range of
rows. It is therefor important to set up the row keys so that related data is
close to each other. For example, if it makes sense to traverse the data in
chronological order, the key might contain a representation of the data's
timestamp. The classic example is to have the reversed domain name (i.e.
com.google.www) as the row key, which means that all sub-domains of any domain
are next to each other (see table 2.1).

@@

BigTable's data model can indeed be thought of as a sparse distributed
multidimensional sorted map. In concept, this multidimensional map is identical
to a nested hash table. The dimensions of this multidimensional map is mapped
out on the row key, the column family and the column qualifier, and possibly
also of the version timestamp.

@@

\begin{lstlisting}[caption=BigTable data model illustration.,captionpos=b]
   (row key,column key,timestamp) -> value
\end{lstlisting}

@@

The data model is a simple one: Rows of data are stored in labeled tables. Each
row is indexed with a row key, a column key, and a timestamp. The columns have
the form "column-family:qualifier", where the column-family is one of a number
of fixed column-families defined by the table's schema, and the qualifier is an
arbitrary string specified by the application. In effect, column-families are
pre-announced when creating the table while qualifiers are not. The contents of
each column family are stored together, so the user will want to store items
that have similar characteristics in the same column family.  Each data point is
called a cell. A cell is an uninterpreted array of bytes, there are no data
types, and it is up to the application to interpret the data correctly. The
table can store any number of timestamped versions of each cell, allowing
versioning of data.

@@

\subsection*{System components}

@@

% storage.
Because the data sources are typically large, the storage needs to be efficient.
Storing the analyzed data can be done in different ways, for instance storing
the raw data or just aggregates. Storing the raw data is very expensive, since
it would basically copy the data to the data store. Because of this the decision
was made to only store aggregates.

There are several different ways of modelling multidimensional data.

TODO: How to discuss modelling of data in a general sense, without mentioning
MapReduce and BigTable?
% thinking in simple data structures,

% TODO: analysis.
- Something about cardinality -

TODO?: When reasoning about dimensions, the issue of cardinality will come
up sooner or later.

The cardinality of a dimension is the number of distinct values of the
dimension, see it as number of members in a set where the dimension is
viewed as a set. For example, the dimension of \textit{logged in} has a
cardinality of two since \textit{logged in} is a boolean variable; the user
is either logged in or not. The \textit{country} dimension has a much
higher cardinality: there are some 300 countries in the world. Certainly,
there could be dimensions whose cardinality is unbound: the \textit{path}
dimension in the web log example, for example, could in theory have a
limitless number of values.

@@

% export.
Due to the several different data sources, there exist several
different types of applications. Therefor the system does not implement any user
interface other than the REST API\footnote{The REST API is detailed in appendix
X}. Making sense of the data is left for the developer.

@@

The four key requirements of the Zohmg system are: ease of use, speed,
scalability, and adoption, where adoption means being able to fit in the
existing software and hardware eco-system.
% dave probably knows a better word than 'fit'.

@@

% scaling.
In order for a computational setup, cluster or other, to be efficient it needs
to scale well, this means adding more hardware should add as close as possible
to the hardware's potential addition to the computational power.

% fit in existing eco system.
The system needs to co-exist and integrate seamlessly with the already existing
hardware and software eco-system at Last.fm. Thus, building on top of the
already existing software a natural choice.

@@

The first section in this chapter will present the work procedure during the
master thesis. The following chapters will discuss selected tools and their
impact on design and implementation choices for the Zohmg system.

@@

% TODO: renskriv.
MapReduce frameworks: Hadoop, disco.

Hadoop is used in production at Last.fm, it made sense to use it because the
work done would integrate with the Last.fm's infrastructure easily.

@@

These databases were explored and tried out. Sadly most of these BigTable-like
databases were in a very early phase of development and hard to even install.
The experience was that HBase was the most mature of them. Maturity was the
compound experience with the software: community, codebase, documentation,
stability etc. HBase is also tightly coupled with Hadoop which is used in
production at Last.fm.

CouchDB was experienced to be mature as well but lacked integration with
MapReduce, all database communication would be necessary to do over HTTP.
Setting up CouchDB with replication required some hands-on, it lacks sorting of
documents which BigTable-like databases utilize.

@@

HBase was benchmarked (the benchmarks was similar to the ones found in the
BigTable paper?).

@@

@@

The first section of this chapter talked about the procedure during the thesis
work. It described the exploratory model and the prototyping performed, as well
as the design phase that led up to the implementation.

@@

@@

A Zohmg project is necessary for using Zohmg; they are created with a management
script, which is also used for interacting with the Zohmg system. When the user
has created a Zohmg project directory further actions can be performed.  The

Zohmg projects reside within a project directory, which contains five
directories: \texttt{clients}, \texttt{config}, \texttt{lib}, \texttt{mappers},
and \texttt{transformers}. The details of these directories are presented
throughout this chapter.

@@

\subsubsection*{Data sets}

@@

A standard MapReduce map function emits key-value pairs. The Zohmg usermap
function, instead, emits triples (tuples with three values), which consist of a
timestamp, a hash table of dimensions and their respective values, and a hash
table with measurements (values). In effect, the dimensions and the timestamp
specify a point in n-space and the measurements represent the observations made
at that point.

@@

The above example log entry is line broken due to space constraints, normally
the entire entry would be on a single line. The example log line contains access
info for a specific request; it contains the key entries: the client's IP
address, timestamp, requested resource, HTTP status code for the request, the
client's user agent etc.

@@

For every line of the log file (the value argument to the map function), a point
in n-space (the dimensions and the timestamp) is emitted together with a count
of pageviews and bytes; one pageview per request and the number of bytes served
in the request. It is possible to add logic here, perhaps only counting requests
from useragents that are not web spiders which resulted in a HTTP 200 status
code.

@@

The transformer middleware application fetches data from HBase with the same
mechanism as the raw data server (via the Thrift interface); it then uses a user
transformer to alter the data and return it to the user. The user transformer is
a Python program in the project's \texttt{transformers} directory. See the
appendix A for more information.

>>>>

Any organization will eventually want to analyze various characteristics of its
service and operations. The services can be anything from web sites to theme
parks; the analysis might include characteristics such as the number of
visitors per day, or the average time a theme park guest spends queueing for a
particular ride. Generally, the services and operations will generate a log for
later analysis. As the services grow in size, so will the data sets they
generate. The size of the data sets at Last.fm is in the range of several
gigabytes per day. Some data sets are many times larger. It is our belief that
this is typical.

An important property of these data sets is that they generally have a regular
structure and are broken down into records that are independent of each other.
The records can be analyzed independently of each other without losing any of
their meaning, as such they are very well suited for parallel analysis.

Each log record typically contains a number of fields. For example, a log of
visits to a web site might contain a time-stamp, information about what country
the visitor originates from, and what web browser used by the user.

When the records in the data set have been analyzed, conclusions can be made
about what countries generate the most traffic to the site, which web browser
is most popular in a certain country, etc. Summaries of data are generally
called aggregates. To aggregate data, for example to compute the sum of
visitors for all European countries, is a common technique for speeding up data
lookup in data stores.

@@

Organizations measure their operations by considering many variables. When
these variables are placed in a spreadsheet, the variables are set on the x and
y axes. Each axis represents a logical grouping of variables. In the example of
a log file of visits to a web site above, country of origin is one such
variable. A variable is generally called a dimension in this thesis.

As an example: The International Aerodyne Corporation, a leading exporter of
woodworking nails, uses a spreadsheet to keep track of the number of nails
exported to each country every month (see table 1.1).

@@

Last.fm is an internet radio station that also has a large focus on metadata.
Their users can submit what music they listen to, this is called
\textit{scrobbling}. Data submitted by users is analyzed and the user is offered
personalized charts, music, and event recommendations, based on their listening.
Users can also edit metadata about artists, albums, events etc; this forms a
musical encyclopedia and community. The Last.fm web site ties together the above
mentioned areas and presents it to the users.

@@

@@

In addition, the secondary goal is to build a tool that allows users to easily
store their data sets in this data store. The tool should provide a simple---as
in minimalistic---mechanism for data analysis (import) and querying (export).

The data analysis will require the user to write a custom import program. The
queries will be performed via an export API for consumption by computer
programs. Common queries must be answered expediently to allow for analysts to
explore data in real time.

The tool will support analytical functions such as drill-down, slicing, etc; it
should also be aware of dimensions in general and the time dimension in
particular. Each dimension contains a measurement in a specified unit.  Example
units are plays, page views, space travels, or such. There should be no upper
bound on either on the number of units nor on the number of dimensions.

\section*{Existing Software}

@@

BigTable is a data store designed for traversal of very large data sets.
BigTable's cells are basically indexed by row key, column key, and a time-stamp.
Each row has a key and any number of cells attached to it. The rows in a table
are sorted by their keys. The typical use-case of BigTable is to scan a range of
rows. It is therefor important to set up the row keys so that related data is
close to each other. For example, if it makes sense to traverse the data in
chronological order, the key might contain a representation of the data's
time-stamp. The classic example is to have the reversed domain name (i.e.
com.google.www) as the row key, which means that all sub-domains of any domain
are next to each other (see table 2.1).

@@

BigTable's data model can indeed be thought of as a sparse distributed
multidimensional sorted map. In concept, this multidimensional map is identical
to a nested hash table. The dimensions of this multidimensional map is mapped
out on the row key, the column family and the column qualifier, and possibly
also of the version time-stamp.

@@

\begin{lstlisting}[caption=BigTable data model illustration.,captionpos=b]
   (row key,column key,time-stamp) -> value
\end{lstlisting}

@@

The data model is a simple one: Rows of data are stored in labeled tables. Each
row is indexed with a row key, a column key, and a time-stamp. The columns have
the form "column-family:qualifier", where the column-family is one of a number
of fixed column-families defined by the table's schema, and the qualifier is an
arbitrary string specified by the application. In effect, column-families are
pre-announced when creating the table while qualifiers are not. The contents of
each column family are stored together, so the user will want to store items
that have similar characteristics in the same column family.  Each data point is
called a cell. A cell is an uninterpreted array of bytes, there are no data
types, and it is up to the application to interpret the data correctly. The
table can store any number of time-stamped versions of each cell, allowing
versioning of data.

@@

\subsection*{System Components}

@@

% import.
The Zohmg import component concists of a framework which enables the user to
write import programs. The import component handles half of the MapReduce
pattern, the reduce part; it also adds an interface for relaying the imported
data to the storage component. The import programs written by the user include
the other half of the MapReduce pattern, namely the map part, and uses the
interface for storing the imported data.

% storage.
The storage component is built upon a BigTable-like database. Although BigTable
is schema less, the user imposes a loose structure by configuring the layout of
the wanted data sets. Basically, this configuration defines the table layout and
defines what structure the imported data must have.

% analysis.
The cardinality of a dimension is the number of distinct values of the
dimension, see it as number of members in a set where the dimension is
viewed as a set. For example, the dimension of \textit{logged in} has a
cardinality of two since \textit{logged in} is a boolean variable; the user
is either logged in or not. The \textit{country} dimension has a much
higher cardinality: there are some 300 countries in the world. Certainly,
there could be dimensions whose cardinality is unbound: the \textit{path}
dimension in the web log example, for example, could in theory have a
limitless number of values.

@@

% export.
Due to the several different data sources, there exist several different types
of applications one could build on top of Zohmg. Therefore the system does not
implement any user interface other than the HTTP API; which takes a query with
time range, unit, and selected units to extract.

When extracting data it is important to remember that data handled by and stored
in Zohmg is not typed in any way. The developer is responsible for interpreting
the data in a correct manner.

@@

The four key requirements of the Zohmg system are: ease of use, speed,
scalability, and adoption, where adoption means being able to fit in the
existing software and hardware ecosystem.

@@

% scaling.
In order for a computational setup, cluster or other, to be efficient it needs
to scale well. In this sense, scaling means that adding more hardware should, in
a best case scenario, add all of the computational power of the new hardware.

% fit in existing eco system.
The system needs to co-exist and integrate seamlessly with the already existing
hardware and software eco-system at Last.fm. Thus, building on top of already
existing software is a natural choice.

@@

% TODO
The first section in this chapter will present the work procedure during the
master thesis. The following sections will discuss selected tools and their
impact on design and implementation choices for the Zohmg system. The last
section will describe the scope of the thesis.

@@

There exists several MapReduce frameworks, for example Apache Hadoop and
Disco\cite{disco} which are both free software. As mentioned in the scope
section Hadoop was selected to be built upon because it already had an
preeminent role in production at Last.fm.

@@

These databases were explored and tested, but most of them were in a very early
phase of development and hard to even install. The experience was that HBase was
the most mature of them. Maturity was the compound experience with the software:
community, codebase, documentation, stability etc. HBase is also tightly coupled
with Hadoop which is used in production at Last.fm.

CouchDB was experienced to be mature as well but lacked integration with
MapReduce, resulting in that all database communication had to be done over
HTTP. This would use vast resources to keep all HTTP connections open, working
around this adds additional work compared to BigTable-like databases. CouchDB
also lacks sorting of documents over a computer cluster, which BigTable-like
databases utilize. The result was that setting up CouchDB with replication,
which would spread the work, required hands-on work.

@@

% TODO: need data on this.
HBase was benchmarked (the benchmarks was similar to the ones found in the
BigTable paper?).

@@

\section{Scope}
% Avgransning.

% TODO:
% * Live here or live in Introduction chapter? (Lots of concepts are
%   presented after the introduction chapter.)
% * Live at an earlier stage in this chapter?
\subsection*{Tools}

The MapReduce framework Hadoop is used in production at Last.fm. Hadoop was
selected to be used for distributed computing, because of the need for simple
integration with already existing infrastructure and developer competence at
Last.fm.

The BigTable-implementation HBase was selected to be used as the underlying data
storage. It was selected because of its integration with Hadoop, and also, out of
the box, solved several Zohmg related distributed storage problems. Several
other distributed databases were tested, but HBase turned out to be the best
alternative because of the easy integration with Last.fm's existing
infrastructure, code maturity, as well as its vibrant community.

\subsection*{Data Store}

The Zohmg data store should just store data, it should have basic import
and export functionality. Zohmg shall not interpret the stored data, this is
left to the user to do in a correct manner. This decision was reached because of
time constraints when implementing Zohmg.

Because BigTable stores data sorted on row keys, it was decided that Zohmg only
supports time-series-based data. Having data sorted between nodes in a computer
cluster balances the network, computational, and storage loads, resulting in
more even utilization of the cluster. Since Zohmg only supports time-series,
each query has to be made over a time range.

It was decided that the row keys should consist of unit and time-stamp. This
would keep related data within close range, e.g. all rows with the same unit
would be stored next to each other. This means reading data with different units
would imply expensive jumps within the stored data. Because of this it was
decided that queries only support a single unit to be queried.

Zohmg is designed as a data store, which simply imports, stores, and exports
data. Although presenting the data is important, this is left to the user to
solve. This restriction was made because of keeping the implementation simple,
and more general.

Because the data sources are typically large, the storage needs to be efficient.
Storing the analyzed data can be done in different ways, for instance storing
the raw data or just aggregates. Storing the raw data is very expensive, since
it would basically copy the data to the data store. Because of this the decision
was made to only store aggregates.

@@

% TODO
The first section of this chapter talked about the procedure during the thesis
work. It described the exploratory model and the prototyping performed, as well
as the design phase that led up to the implementation.

@@

The last section describes the scope of the thesis.

@@

Zohmg uses the concept of projects. These projects are self contained
application data directories which are used with Zohmg. A Zohmg project is
necessary for using Zohmg; they are created with a management script, which is
also used for interacting with the Zohmg system. When the user has created a
Zohmg project directory further actions can be performed. The Zohmg projects
reside within a project directory, which contains five directories:
\texttt{clients}, \texttt{config}, \texttt{lib}, \texttt{mappers}, and
\texttt{transformers}. The details of these directories are presented throughout
this chapter.

@@

\subsubsection*{Data Sets}

@@

A standard MapReduce map function emits key-value pairs. The Zohmg usermap
function, instead, emits triples (tuples with three values), which consist of a
time-stamp, a hash table of dimensions and their respective values, and a hash
table with measurements (values). In effect, the dimensions and the time-stamp
specify a point in n-space and the measurements represent the observations made
at that point.

@@

The above example log entry is line broken due to space constraints, normally
the entire entry would be on a single line. The example log line contains access
info for a specific request; it contains the key entries: the client's IP
address, time-stamp, requested resource, HTTP status code for the request, the
client's user agent etc.

@@

For every line of the log file (the value argument to the map function), a point
in n-space (the dimensions and the time-stamp) is emitted together with a count
of pageviews and bytes; one pageview per request and the number of bytes served
in the request. It is possible to add logic here, perhaps only counting requests
from useragents that are not web spiders which resulted in a HTTP 200 status
code.

@@

The transformer middleware application fetches data from HBase with the same
mechanism as the raw data server (via the Thrift interface); it then uses a user
transformer to alter the data and return it to the user. The user transformer is
a Python program in the project's \texttt{transformers} directory. See appendix
A for more information.
